### 2022-12-27 paper.Rmd
minor

Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---which seeks homogeneity within clusters and separation between clusters [@rokach2005; @steinley2006]. Anticlustering has many applications in research psychology [@steinley2006; @brusco2019; @papenberg2020]: Examples include splitting tests into parts of equal difficulty [@gierl2017], assigning students to work groups [@baker2002], and assigning stimuli to different, but parallel, experimental conditions [@lahl2006]. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, these problems can be formalized as mathematical optimization problems [e.g., @brusco2019; @spath1986; @baker2002; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods are characterized by (a) an objective function that quantifies between-group similarity and/or within-group heterogeneity and (b) an algorithm that determines how elements are assigned to groups in such a way that the objective function is maximized. Several anticlustering objective functions use pairwise dissimilarity ratings such as the Euclidean distance as input. The most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between any elements within the same group [@brusco2019; @gallego2013]. By summing all pairwise dissimilarities of elements within the same group, the diversity reflects the total degree of within-group heterogeneity. High within-group diversity simultaneously ensures high similarity of the different groups.[^equalsizedgroups] Another important anticlustering criterion is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications requiring high within-group heterogeneity should incorporate both dispersion and diversity, presenting a bicriterion algorithm to approximate the Pareto efficient set of solutions according to both criteria.

[^equalsizedgroups]: As will be discussed later, the diversity objective only stricly equalizes within-group heterogeneity and between-group similarity when groups are equal-sized.

Oftentimes, partitioning applications in psychology are focused on ensuring between-group similarity rather than within-group heterogeneity (even though both goals oftentimes coincide). High similarity between-group similarity is for example desireable when designing experimental conditions that have to incorporate different stimulus sets that should however be as similar as possible according to response-relevant attributes [@lahl2006]. Such stimulus assignment tasks are usually conducted on the basis of attribute data [or *features*; @dry2009] rather than pairwise dissimilarity ratings, for example when selecting stimuli for an experiment on the basis of norming data [e.g., @kurdi2017introducing]. For word stimuli in psycholinguistic experiments, norming data may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attribute values are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, an anticlustering criterion such as the diversity can be optimized to obtain a partitioning of the stimuli. A different anticlusting approach employs clustering objectives that can directly be computed on the attribute values. K-means is probably the best-known objective of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other parameters---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize similarity with regard to the mean; even an entirely random group assignment usually yielded more similar variances. When requiring overall between-group similarity, the negligence of other distribution characteristics than the mean is misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

@@

This paper's focus on the k-plus objective only instead of optimization algorithms followed two reasonings. First, the applications showed that the local maximum search---maximizing the unweighted k-plus sum---yields satisfying results in practice that probably suffice in most applications. Second, it was of interest to investigate the merits of the k-plus objective function per se, independently from the algorithm that is used to optimize it. To obtain a fair assessment In the simulation study, it was therefore important that each anticlustering objective is optimized using the same algorithm, instead of using specialized optimization schemes tailored to each individual criterion. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering.

>>>>

Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity [@brusco2019; @spath1986; @valev1998]. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---which seeks homogeneity within clusters and separation between clusters [@rokach2005; @steinley2006]. Anticlustering has many applications in research psychology [@steinley2006; @brusco2019; @papenberg2020]. Examples include splitting tests into parts of equal difficulty [@gierl2017], assigning students to work groups [@baker2002], and assigning stimuli to different, but parallel, experimental conditions [@lahl2006]. Solving anticlustering problems "by hand" is a tedious and time-consuming task, and the quality of manual partitioning is usually subpar. Fortunately, anticlustering problems can be formalized as mathematical optimization problems [e.g., @brusco2019; @spath1986; @baker2002; @fernandez2013] and accessible open source software solutions to tackling these problems exist [@papenberg2020].

Anticlustering methods are characterized by (a) an objective function that quantifies between-group similarity and/or within-group heterogeneity and (b) an algorithm that determines how elements are assigned to groups to maximize the objective function Several anticlustering objective functions use pairwise dissimilarity ratings such as the Euclidean distance as input. The most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between any elements within the same group [@brusco2019; @gallego2013]. By summing all pairwise dissimilarities of elements within the same group, the diversity reflects the total degree of within-group heterogeneity. High within-group diversity simultaneously ensures high between-group similarity.[^equalsizedgroups] Another important anticlustering criterion is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases the within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications requiring high within-group heterogeneity should incorporate both dispersion and diversity, and presented a bicriterion algorithm to approximate the Pareto efficient set of solutions according to both criteria.

[^equalsizedgroups]: As will be shown later, the diversity objective only strictly equalizes within-group heterogeneity and between-group similarity when groups are equal-sized.

Oftentimes, partitioning applications in psychology are focused on ensuring between-group similarity rather than within-group heterogeneity (even though both goals usually coincide). High similarity between-group similarity is for example desirable when preparing experimental conditions that use different stimulus sets that should however be as similar as possible with regard to response-relevant attributes [@lahl2006]. Rather than relying on pairwise (dis)similarity ratings, stimulus selection tasks are usually conducted on the basis of attribute data [or *features*; @dry2009]. For stimuli in psycholinguistic experiments, for example, attributes may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attributes are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, a distance-based anticlustering criterion such as the diversity can be optimized to obtain a partitioning of the stimuli. A different anticlusting approach employs clustering objectives that can directly be computed on the attribute values. K-means is probably the best-known objective of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized [@brusco2006branch], usually using the k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other parameters---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize similarity with regard to the mean; even an entirely random group assignment usually yielded more similar variances. When requiring overall between-group similarity, the negligence of other distribution characteristics than the mean is misguided, because similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

@@

This paper's focus on the k-plus objective only instead of optimization algorithms followed two reasonings. First, the applications showed that the local maximum search---maximizing the unweighted k-plus sum---yields satisfying results in practice that probably suffice in most applications. Second, it was of interest to investigate the merits of the k-plus objective function per se, independently from the algorithm that is used to optimize it. To obtain a fair assessment In the simulation study, it was therefore important that each anticlustering objective was optimized using the same algorithm. Differences in performance cannot be interpreted if the criterion as well as the algorithm differs between methods. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering.
